Immutable Source-Mapped Abstract Syntax Tree:\ \\ A Design Pattern for Refactoring Engine APIs

### 2016-07-14 paper.tex
Variants are now discussed more explicity.

% Title portion
\title{Immutable Source-Mapped Abstract Syntax Tree:\ \\
A Design Pattern for Refactoring Engine APIs} % \\ \today}
\author{JEFFREY L.\ OVERBEY and CARL R.\ WORLEY \affil{Auburn University}}

@@

When ASTs are used in compilers, it is common to make them \textit{mutable},
meaning that nodes can easily be added, changed, or deleted from the tree.  For
example, a compiler might determine that \ttt{a~=~b*1} can be simplified to
\ttt{a~=~b}, and then change the AST accordingly.

@@

Unfortunately, while a mutable AST can provide an elegant means of manipulating
source code, that are other issues that can make mutable ASTs problematic from
an API design perspective.  Some notable issues include the following.
\begin{itemize}
\item \textit{AST nodes contain source location information.}  After the
structure of the AST is changed, what does the source location information in
each node represent: the ``old'' source location (in the original code) or the
``new'' location (after it has been modified)?  If it represents the old
location, what should the source location be when new nodes are inserted into
the tree?  If it represents the new location, then every time a change is made
to the tree, source location information must be updated for every node in the
tree, which may be expensive \dots or else a clever implementation (e.g.,
computation from relative offsets) must be used instead.
\item \textit{AST nodes may be removed from the tree, leaving them orphaned.}
Suppose a client traverses the AST, storing pointers to some ``interesting''
nodes in a collection.  Then, an ancestor of one of those nodes is removed from
the AST.  This will leave the client in a difficult-to-debug situation where
his collection contains pointers to nodes that no longer exist in the AST.
\item \textit{Semantic information may be invalidated.}  Refactorings often
need to determine, for a particular use of a variable, what declaration that
use corresponds to.  For example, in Java, if a field and a local variable were
both named \ttt{x}, then in the expression \ttt{this.x = x}, the first \ttt{x}
would refer to the field, while the second would refer to the local variable.
However, if the local variable declaration were removed, the second \ttt{x}
would refer to the field instead.  This can be confusing to clients, since
information collected about the program at one point in time may no longer be
true after the tree's structure changes.
\item \textit{It may be possible to mutate the AST so that it does not
represent a legal program.}  It might even be possible to mutate the AST so
that it is no longer a tree at all.  In Java, classes can contain other
classes.  What happens if a client tries to make a class an inner class of
itself?  (This could turn the tree into a cyclic graph.)  The API for modifying
the AST's structure could try to prevent such things, but attempting to detect
and prevent every possible mistake would be difficult and potentially
expensive.
\item \textit{Mutability can be problematic in multithreaded code.}
\end{itemize}

@@

Guidelines for API design are discussed in detail in Jaroslav Tulach's
\textit{Practical API Design}~\cite{tulach08practical}.  A presentation by
Joshua Bloch~\cite{bloch} offers excellent advice as well.  Both authors
suggest minimizing the amount of mutability in an API.  Immutable objects tend
to have simpler interfaces, and they are thread-safe.  Both authors also
emphasize that APIs should be easy to learn (even for programmers who do not
read all of the API documentation) and difficult for clients to misuse.  Given
these guidelines, and the list of liabilities for mutable ASTs given above, it
seems preferable from the perspective of someone attempting to publish an API
to avoid mutable ASTs.  But this raises a different question: If ASTs are
immutable, what API should be provided for clients to manipulate source code?

@@

So, is it possible to provide a cleaner source manipulation API, while still
using an immutable AST?  It is, but some additional API is needed.

One option is to provide an API that allows clients to specify what changes
they want to make in terms of AST nodes.  This API can then translate these
into string edits.  For example, an API might offer a \textit{Rewriter} class.
The client would inform the \textit{Rewriter} that it wants to effectively
delete a particular \textit{FieldDeclaration} node from the AST:
\textit{rewriter.delete(field)}.  When it is ready to manipulate the source
code, it delegates this task to the rewriter:
\textit{rewriter.rewrite(sourceCode)}.  This is illustrated in
Figure~\ref{fig:seq-diagram}.  The Eclipse Java and C/C++ Development
Tools~\cite{cdt-refactoring} take this approach (with slightly different class
and method names).

@@

A second option is to work with ASTs in a functional style, where every edit
returns a new AST.  So, a call to \textit{field.removeFromTree()} would not
change the existing AST, but rather it would return a new AST that has the
given field omitted.  (Of course, this requires a very careful implementation.
Memory consumption and copying are a concern, particularly when ASTs are large.
Also, a mechanism must be provided so that clients can easily map nodes in the
original tree to the ``same'' node in the modified tree, and vice versa.)

>>>>

% Title portion
\title{Immutable Source-Mapped Abstract Syntax Tree:\ \\
A Design Pattern for Refactoring Engine APIs} % \\ \today}
\author{JEFFREY L.\ OVERBEY \and CARL R.\ WORLEY \affil{Auburn University}}

@@

There are two main variants to the Abstract Syntax Tree pattern for refactoring:
mutable and immutable. A mutable tree has nodes that can be easily added, 
changed, or deleted from the tree. These trees are often used internally by
compilers. For example, a compiler might determine that
\ttt{a~=~b*1} can be simplified to \ttt{a~=~b}, and then change the AST 
accordingly.

@@

Mutable ASTs are conceptually simple and allow code for specific refactorings
to be very short. Performing refactorings on source code is conceptually
identical to changing the AST itself. This can greatly lower the bar for 
designing refactorings. However, mutable ASTs introduce a host of problems when
used in a large API:

\begin{itemize}
\item \textit{AST nodes contain source location information.}  After the
structure of the AST is changed, what does the source location information in
each node represent: the ``old'' source location (in the original code) or the
``new'' location (after it has been modified)?  If it represents the old
location, what should the source location be when new nodes are inserted into
the tree?  If it represents the new location, then every time a change is made
to the tree, source location information must be updated for every node in the
tree, which may be expensive \dots or else a clever implementation (e.g.,
computation from relative offsets) must be used instead.
\item \textit{AST nodes may be removed from the tree, leaving them orphaned.}
Suppose a client traverses the AST, storing pointers to some ``interesting''
nodes in a collection.  Then, an ancestor of one of those nodes is removed from
the AST.  This will leave the client in a difficult-to-debug situation where
his collection contains pointers to nodes that no longer exist in the AST.
\item \textit{Semantic information may be invalidated.}  Refactorings often
need to determine, for a particular use of a variable, what declaration that
use corresponds to.  For example, in Java, if a field and a local variable were
both named \ttt{x}, then in the expression \ttt{this.x = x}, the first \ttt{x}
would refer to the field, while the second would refer to the local variable.
However, if the local variable declaration were removed, the second \ttt{x}
would refer to the field instead.  This can be confusing to clients, since
information collected about the program at one point in time may no longer be
true after the tree's structure changes.
\item \textit{It may be possible to mutate the AST so that it does not
represent a legal program.}  It might even be possible to mutate the AST so
that it is no longer a tree at all.  In Java, classes can contain other
classes.  What happens if a client tries to make a class an inner class of
itself?  (This could turn the tree into a cyclic graph.)  The API for modifying
the AST's structure could try to prevent such things, but attempting to detect
and prevent every possible mistake would be difficult and potentially
expensive.
\item \textit{Mutability can be problematic in multithreaded code.}
\end{itemize}

@@

Guidelines for API design are discussed in detail in Jaroslav Tulach's
\textit{Practical API Design}~\cite{tulach08practical}.  A presentation by
Joshua Bloch~\cite{bloch} offers excellent advice as well.  Both authors
suggest minimizing the amount of mutability in an API.  Immutable objects tend
to have simpler interfaces, and they are thread-safe.  Both authors also
emphasize that APIs should be easy to learn (even for programmers who do not
read all of the API documentation) and difficult for clients to misuse. Mutable
ASTs have the advantage of ease-of-learning, but have have all the disadvantages
listed above. They can be valid solutions for simpler refactorings or projects
developed by a small team of people, but become problematic when used for 
nontrivial refactorings or for a public API. For a general, public refactoring
API, immutable trees are less error-prone.

@@

So, is it possible to provide a cleaner source manipulation API, while still
using an immutable AST?  It is, and there are two main variants that allow
for this.

\subsection{Variants}

    \item \textit{Work with the ASTs in a functional style, where every edit
        returns a new AST.} So, a call to field.removeFromTree() would not
        change the existing AST, but rather it would return a new AST that has
        the given field omitted. When dealing with immutable data structures,
        functional programming may often be the natural solution. However,
        in the case of ASTs, a functional style risks reintroducing many of 
        the problems inherent to mutable trees. File location of nodes may
        be disrupted, pointers to nodes in the original tree may be rendered 
        irrelevant, semantic information may be invalidated, etc. Those are in
        addition to the difficulties of functional programming, such as memory
        consumption and copying, especially with very large ASTs. This is
        a valid refactoring API design pattern, but any implementation using 
        this variant must carefully consider these problems.

    \item \textit{Provide an API that allows clients to specify what changes
        they want to make in terms of AST nodes.}  This API can then translate these into string edits.  For example, an API might offer a \textit{Rewriter} 
class. The client would inform the \textit{Rewriter} that it wants to 
effectively delete a particular \textit{FieldDeclaration} node from the AST:
\textit{rewriter.delete(field)}.  When it is ready to manipulate the source
code, it delegates this task to the rewriter:
\textit{rewriter.rewrite(sourceCode)}.  This is illustrated in
Figure~\ref{fig:seq-diagram}.  The Eclipse Java and C/C++ Development
Tools~\cite{cdt-refactoring} take this approach (with slightly different class
and method names).

@@
